Graph Topological Aspects of Granger Causal Network Learning\\ \large Supplementary Material

### 2019-08-27 paper_theory/supplementary_material.tex
some preliminary analysis on GC appl paper.

\begin{lemma}
  \label{lem:adj_matrix}
  Let $S$ be the transposed adjacency matrix\footnote{\footnotesize We
    are using the convention that $\B_{ij}(z)$ is a filter with input
    $x_j$ and output $x_i$ so as to write the action of the system as
    $\B(z)x(t)$ with $x(t)$ as a column vector.  This competes with
    the usual convention for adjacency matrices where $A_{ij} = 1$ if
    there is an edge $(i, j)$.  In our case, the sparsity pattern of
    $\B_{ij}$ is the \textit{transposed} conventional adjacency
    matrix.} of the Granger-causality graph $\gcg$.  Then,
  $(S^k)_{ij}$ is the number of paths of length $k$ from node $j$ to
  node $i$.  Evidently, if $\forall k \in \N,\ (S^k)_{ij} = 0$ then
  $j \not\in \anc{i}$.
\end{lemma}
\begin{proof}
  This is a well known theorem, proof follows by induction.
\end{proof}

@@

\begin{proposition}
  \label{prop:separated_ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  If

@@

\begin{lemma}
  \label{lem:vj_perp}
  Consider distinct nodes $i, j$ in a Granger-causality graph $\gcg$.
  If $j \not \in \anc{i}$, then $\H_t^{(v_j)} \perp \H_{t}^{(i)}$, and therefore
  for any causal filter $\Phi(z)$ we have

@@

\begin{proposition}
  \label{prop:ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph $\gcg$.
  If

@@

  % \BlankLine
  \caption{Pairwise Granger Causality Algorithm (PWGC)}
  \label{alg:pwgr}
  % \TitleOfAlgo{Pairwise Graph Recovery (PWGC)}
  % \Input{Pairwise Granger-causality relations between a persistent
  % process of dimension $n$ whose joint Granger-causality
  % relations are known to form a strongly causal DAG $\gcg$.}
  \Input{Pairwise Granger-causality relations}
  \Output{Edges $\gcge = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of
    the graph $\gcg$.}
  \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
    $E_0 = \emptyset$  \texttt{\# edges of }$\gcg$\\
    % $P_0 = \emptyset$  \texttt{\# layer by layer driving nodes}\\
    $k = 1$ \texttt{\# a counter used only for notation}}
  \BlankLine
  $W \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# candidate edges}\\
  $P_0 \leftarrow \{i \in S_0\ |\ \forall s \in S_0\ (s, i) \not\in W\}$  \texttt{\# parentless nodes}\\
  \While{$S_{k - 1} \ne \emptyset$}{
    $S_k \leftarrow S_{k - 1} \setminus P_{k - 1}$ \texttt{\# remove nodes with depth }$k - 1$\\
    $P_k \leftarrow \{i \in S_k\ |\ \forall s \in S_k\ (s, i) \not\in W\}$   \texttt{\# candidate children}\\

@@

In subsection \ref{sec:pairwise_hypothesis_testing} we define pairwise
Granger-causality hypothesis tests, in subsection
\ref{sec:model_order_selection} a model order selection criteria, in
subsection \ref{sec:efficient_model_estimation} an efficient
estimation algorithm, in subsection \ref{sec:error_rate_control} the
method for choosing an hypothesis testing threshold, and finally in
subsection \ref{sec:finite_pwgc} the unified finite sample algorithm.

\subsection{Pairwise Hypothesis Testing}
\label{sec:pairwise_hypothesis_testing}
In performing pairwise checks for Granger-causality $x_j \pwgc x_i$ we
follow the simple scheme of estimating the following two linear models:

@@

\subsection{Edge Probabilities and Error Rate Controls}
\label{sec:error_rate_control}
Denote $F_{ij}$ the Granger-causality statistic of Equation
\eqref{eqn:gc_statistics} with model orders chosen by the methods of
Section \ref{sec:model_order_selection}.  We assume that this
statistic is asymptotically $\chi^2(p_{ij})$ distributed (e.g. the
disturbances are Gaussian), and denote by $G$ the cumulative
distribution function thereof.  We will define the matrix

@@

    \Input{Estimates of pairwise Granger-causality statistics $F_{ij}$
      (eqn. \ref{eqn:gc_statistics}).  Matrix of edge probabilities $P_{ij}$ (eqn. \ref{eqn:edge_inclusion_probability}).  Hypothesis testing threshold $\delta$ chosen via the Benjamini-Hochberg criterion (Section \ref{sec:error_rate_control})}
    \Output{A strongly causal graph $\widehat{\gcg}$}
    \Initialize{$S = [n]$  \texttt{\# unprocessed nodes}\\
      $E = \emptyset$  \texttt{\# edges of }$\widehat{\gcg}$\\
      $k = 1$ \texttt{\# a counter used only for notation}}

@@

In regards scalability, we have observed that performing the $O(n^2)$
pairwise Granger-causality calculations consumes the vast majority
($> 90\%$) of the computation time.  Since this step is trivially
parallelizable, our algorithm also scales well with multiple cores or
multiple machines.  Figure \ref{fig:simulation_results_scaling} is a
demonstration of this scalability, where we are able to estimate
graphs having over $1500$ nodes (over $2.25 \times 10 ^6$ possible edges)
using only $T = 500$ data points, granted, an SCG on this many nodes
is extremely sparse.

>>>>

\begin{lemma}
  \label{lem:adj_matrix}
  Let $S$ be the transposed adjacency matrix\footnote{\footnotesize We
    are using the convention that $\B_{ij}(z)$ is a filter with input
    $x_j$ and output $x_i$ so as to write the action of the system as
    $\B(z)x(t)$ with $x(t)$ as a column vector.  This competes with
    the usual convention for adjacency matrices where $A_{ij} = 1$ if
    there is an edge $(i, j)$.  In our case, the sparsity pattern of
    $\B_{ij}$ is the \textit{transposed} conventional adjacency
    matrix.} of the Granger causality graph $\gcg$.  Then,
  $(S^k)_{ij}$ is the number of paths of length $k$ from node $j$ to
  node $i$.  Evidently, if $\forall k \in \N,\ (S^k)_{ij} = 0$ then
  $j \not\in \anc{i}$.
\end{lemma}
\begin{proof}
  This is a well known theorem, proof follows by induction.
\end{proof}

@@

\begin{proposition}
  \label{prop:separated_ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger causality graph
  $\gcg$.  If

@@

\begin{lemma}
  \label{lem:vj_perp}
  Consider distinct nodes $i, j$ in a Granger causality graph $\gcg$.
  If $j \not \in \anc{i}$, then $\H_t^{(v_j)} \perp \H_{t}^{(i)}$, and therefore
  for any causal filter $\Phi(z)$ we have

@@

\begin{proposition}
  \label{prop:ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger causality graph $\gcg$.
  If

@@

  % \BlankLine
  \caption{Pairwise Granger Causality Algorithm (PWGC)}
  \label{alg:pwgr}
  % \TitleOfAlgo{Pairwise Graph Recovery (PWGC)}
  % \Input{Pairwise Granger causality relations between a persistent
  % process of dimension $n$ whose joint Granger causality
  % relations are known to form a strongly causal DAG $\gcg$.}
  \Input{Pairwise Granger causality relations}
  \Output{Edges $\gcge = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of
    the graph $\gcg$.}
  \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
    $E_0 = \emptyset$  \texttt{\# edges of }$\gcg$\\
    % $P_0 = \emptyset$  \texttt{\# layer by layer driving nodes}\\
    $k = 1$ \texttt{\# a counter used only for notation}}
  \BlankLine
  $W \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# candidate edges}\\
  $P_0 \leftarrow \{i \in S_0\ |\ \forall s \in S_0\ (s, i) \not\in W\}$  \texttt{\# parentless nodes}\\
  \While{$S_{k - 1} \ne \emptyset$}{
    $S_k \leftarrow S_{k - 1} \setminus P_{k - 1}$ \texttt{\# remove nodes with depth }$k - 1$\\
    $P_k \leftarrow \{i \in S_k\ |\ \forall s \in S_k\ (s, i) \not\in W\}$   \texttt{\# candidate children}\\

@@

In subsection \ref{sec:pairwise_hypothesis_testing} we define pairwise
Granger causality hypothesis tests, in subsection
\ref{sec:model_order_selection} a model order selection criteria, in
subsection \ref{sec:efficient_model_estimation} an efficient
estimation algorithm, in subsection \ref{sec:error_rate_control} the
method for choosing an hypothesis testing threshold, and finally in
subsection \ref{sec:finite_pwgc} the unified finite sample algorithm.

\subsection{Pairwise Hypothesis Testing}
\label{sec:pairwise_hypothesis_testing}
In performing pairwise checks for Granger causality $x_j \pwgc x_i$ we
follow the simple scheme of estimating the following two linear models:

@@

\subsection{Edge Probabilities and Error Rate Controls}
\label{sec:error_rate_control}
Denote $F_{ij}$ the Granger causality statistic of Equation
\eqref{eqn:gc_statistics} with model orders chosen by the methods of
Section \ref{sec:model_order_selection}.  We assume that this
statistic is asymptotically $\chi^2(p_{ij})$ distributed (e.g. the
disturbances are Gaussian), and denote by $G$ the cumulative
distribution function thereof.  We will define the matrix

@@

    \Input{Estimates of pairwise Granger causality statistics $F_{ij}$
      (eqn. \ref{eqn:gc_statistics}).  Matrix of edge probabilities $P_{ij}$ (eqn. \ref{eqn:edge_inclusion_probability}).  Hypothesis testing threshold $\delta$ chosen via the Benjamini-Hochberg criterion (Section \ref{sec:error_rate_control})}
    \Output{A strongly causal graph $\widehat{\gcg}$}
    \Initialize{$S = [n]$  \texttt{\# unprocessed nodes}\\
      $E = \emptyset$  \texttt{\# edges of }$\widehat{\gcg}$\\
      $k = 1$ \texttt{\# a counter used only for notation}}

@@

In regards scalability, we have observed that performing the $O(n^2)$
pairwise Granger causality calculations consumes the vast majority
($> 90\%$) of the computation time.  Since this step is trivially
parallelizable, our algorithm also scales well with multiple cores or
multiple machines.  Figure \ref{fig:simulation_results_scaling} is a
demonstration of this scalability, where we are able to estimate
graphs having over $1500$ nodes (over $2.25 \times 10 ^6$ possible edges)
using only $T = 500$ data points, granted, an SCG on this many nodes
is extremely sparse.
